Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-01-16 Manuscript/manuscript.Rmd
manuscript edits

Replicate Hi-C data for the GM12878 cell line were downloaded from GEO GSE63525 (Experiments HIC001 to HIC029) [@rao20143d]. The chromosomal coordinates of topologically associating domains throughout the genome were obtained using the Arrowhead algorithm from the Juicer toolbox provided by the Aiden Laboratory (https://github.com/aidenlab/juicer) [@durand2016juicebox]. [??? how many; !!! the first 29 hic matrices (HIC001 to HIC029); should I provide more detail? ??? TBD, why 29?] contact matrices at 5 kb, 25 kb, 50 kb, and 100 kb resolutions. Identified TAD boundaries were represented by their location on the linear genome (GRCh37/hg19 human genome assembly), including start and end coordinates for chromosomes 1 through 22. The start and end coordinates of all hierarchical TADs were concatenated, sorted, and unique TAD boundary points were obtained.

@@

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution (Figure 1). For example, for a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. To accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sized by regions of size of the resolution of the corresponding contact matrix. For example, for the domain data obtained from a 10 kb resolution contact matrix, each uniquely identified TAD boundary was flanked on either side by 10 kb for a total width around each boundary point of 20 kb. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

@@

Genomic annotation data (GRCh37/hg19 human genome assembly) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 1). The annotations consisted of a variety of different chromatin states, histone marks, and DNase I hypersensitive sites.

@@

Each genomic annotation is typically represented by a set of regions (their genomic coordinates) annotated as having a functional property or biological annotation. Using these sets of genomic annotations, we created three types of predictors that were used as features to analyze their relationships with TAD boundaries (Figure 2):

@@

3. **Distance**: For each genomic bin, the distance (in bases) from the center of the bin to the center of the nearest region of a respective genomic annotation region was calculated.

An ensemble of models was established for the purposes of systematically comparing the performances of different parameters in the framework. Each system in the framework is assumed to be composed of a model constructed from a particular data resolution, with a particular predictor type for the feature space (Figure 3).

The data sets used for modeling were composed of the full set of genomic bins that either overlapped with a flanked TAD boundary or did not. These two classes made up the majority and minority sets. The data was then split into a 7:3 ratio of training to testing sets. Each training and testing set was composed of a similar majority to minority class ratio as the full data set. For each training set, an $l_{1}$ norm penalized regularization (LASSO) was performed [??? what for?]. The $\lambda$ penalization term was tuned over a grid of 10 values on an exponential scale ranging from 0.01 to 10.

@@

## Figure 2: Predictor Types

@@

## Figure 3: Ensemble Framework/Model Building Pipeline

>>>>

Replicate Hi-C data for the GM12878 cell line were downloaded from GEO GSE63525 (Experiments HIC001 to HIC018; produced using the MbolI restriction enzyme) [@rao20143d]. The chromosomal coordinates of topologically associating domains throughout the genome were obtained using the Arrowhead algorithm from the Juicer toolbox provided by the Aiden Laboratory (https://github.com/aidenlab/juicer) [@durand2016juicebox]. Domain data was extracted at 4 different kilobase (kb) resolutions including 10 kb, 25 kb, 50 kb, and 100 kb. Identified TAD boundaries were represented by their location on the linear genome (GRCh37/hg19 human genome assembly), including start and end coordinates for chromosomes 1 through 22. The start and end coordinates of all hierarchical TADs were concatenated, sorted, and unique TAD boundary points were obtained.

@@

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution (Figure 1). For example, fordomain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. To accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. For example, returning to the 10 kb resolution example, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

@@

Genomic annotation data (GRCh37/hg19 human genome assembly) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 1). The annotations consisted of a variety of different chromatin states, histone marks, and transcription factor binding sites using ChIP-seq peak experiments. Each genomic annotation was represented by a set of regions (their genomic coordinates) annotated as having a functional property or biological significance.

An ensemble of models was established for the purposes of systematically comparing the performances of different parameters in the framework. Each system in the framework is assumed to be composed of a model constructed from a particular data resolution, with a particular predictor type for the feature space (Figure 2).

The data sets used for modeling were composed of the full set of genomic bins that either overlapped with a flanked TAD boundary or did not. These two classes made up the minority and majority sets respectively. The data was then split into a 7:3 ratio of training to testing sets. Each training and testing set was composed of a similar majority to minority class ratio as the full data set. That is, for the task of prediction, the test set remained constant for all ensemble systems in order to account for a fair and unbiased comparison.

@@

Using the sets of genomic annotations, we created three types of predictors that were used as features to analyze their relationships with TAD boundaries (Figure 3):

@@

3. **Log Distance**: For each genomic bin, the log of the distance (in bases) from the center of the bin to the center of the nearest region of a respective genomic annotation region was calculated. The log transformation was used as a normalization technique in order to account for the skewness of the data (Supplementary Figure 1)

To assess the predictive performance for each predictor type, we performed individual random forest models utilizing each specific predictor type, in addition to a model with all combined types. Precision-Recall (PR) Curves, and subsequent area under the PR curves (AUPRC), were used to evaluate which predictor type contributed to the best model performance.

### VARIABLE REDUCTION

@@

## Figure 2: Ensemble Framework/Model Building Pipeline

## Figure 3: Predictor Types

@@
